Analysis of the 2005--\myrevision{2016} Earthquake Sequence in Northern Iran Using the Visibility Graph Method

### 2017-06-14 seismicity.tex
First revision draft for PAAG.

We are interested in applying the visibility graph method to analyze the seismicity of northern Iran. This region is part of the Iranian plateau, on the Himalayan-Alpine seismic belt. It is confined by the relative movements between the Arabian, Eurasian, eastern Asia-Minor, and Indian plates; and has a long history of large magnitude ($M>7$) earthquakes that are well documented dating back to the eight century \citep[e.g.,][]{Berberian_1981_Chap}. According to the tectonic settings and geologic provinces of the plateau, the seismic activity in Iran has been categorized into different seismic zones. These vary between four to nine major seismic zones, in the more traditional models \citep[e.g.,][]{Stocklin1968, Takin1972, Berberian1976}, up to twenty to twenty-three seismotectonic provinces in the most elaborate ones \citep[e.g.,][]{Nowroozi1976, Tavakoli1999}. We adopt the model proposed by \citet{Mirzaei1998} with modifications introduced by \citet{Karimiparidari2013}. In it, Iran is divided into six seismic regions: Azerbaijan, Alborz, Kopeh Dagh, Zagros, Central-East Iran, and Makran. Our focus, however, is on the northern part of Iran, for which we define a region of interest between longitudes 43.5\textdegree{}E and 61.5\textdegree{}E, and latitudes 34\textdegree{}N and 40\textdegree{}N, as shown in Fig.~\ref{fig:study_region}. Although this region encloses part of the Zagros and Central-East seismic zones, we concentrate in the analysis of the seismicity of Azerbaijan, Alborz and Kopeh Dagh.

@@

\begin{figure*}[t]
	\centering
	\includegraphics[width=\textwidth]{figures/pdf/figure-03} 
	\caption{Declustered instrumental seismicity of northern Iran for the 2005--2015 period, considering only the events in the seismic zones of interest, namely Azerbaila, Alborz, and Kopeh Dagh. The size of symbols are proportional to the magnitude of the events, as indicated by the artificial scale shown on the right. The background shows fault lines and the shaded relief. The color version of this figure is available only in the electronic edition.}
	\label{fig:seismicity}
\end{figure*}

It follows from this description that the region of interest is one of significant seismic activity, which goes back centuries. We are, however, interested in the most recent instrumental seismicity. In particular, we focus on events recorded in the last decade, between January 2005 and December 2015. We focus our analysis on this magnitude-time sequence period because in a previous study we observed that the latest decade of seismic records in northern Iran offered the most complete dataset of recorded events, especially for the case of small $M < 4$ earthquakes \citep[e.g.][]{Khoshnevis2016}.

We compiled a catalog of all recorded earthquakes using data downloaded from the International Institute of Earthquake Engineering and Seismology, IIEES (see the Data and Resources section). The obtained dataset contained a mixture of earthquake magnitude scales, including: moment, $M_w$; local, $M_L$; body wave, $m_b$; surface wave, $M_s$; and duration $M_D$ magnitudes. We converted all earthquakes to moment magnitude, $M_w$. Magnitudes in the $M_L$, $m_b$, and $M_s$ scales were converted using the relationships defined in \citet{Zare2014}, whereas $M_D$ magnitudes were converted using the empirical relationship proposed by \citet{Deniz2010}.

Foreshocks and aftershocks were removed from the catalog following a Poissonian occurrence model using the declustering method of \citet{Gardner1974}, in a manner consistent with previous catalog compilations available for the region \citep[e.g.,][]{Zare2014}. Upon declustering, we divide the catalog into three sub-catalogs, each for one of the seismic zones under consideration. Fig.~\ref{fig:seismicity} shows the epicenter location of declustered instrumental earthquakes for the three seismic zones of interest.

We then proceeded to determine the minimum magnitude of completeness, $M_c$, for each region's sub-catalog. The value of $M_c$ is often determined using simple numerical analyses in combination with data inspection. Two common approaches are the maximum curvature (MAXC) and the goodness-of-fit test (GFT) methods \citep{Wiemer2001}. We use the GFT method as introduced by \citet{Wiemer2000}. In this method, the completeness magnitude is obtained such that the catalog satisfies---at a certain acceptance threshold---the Gutenberg-Richter power low, given an extended dataset of events. This extended dataset is composed of predicted and observed events, where the predicted events are generated based on a trial minimum magnitude. The process is repeated for increasing values of the reference minimum magnitude until finding a desirable fit with the power law. \citet{Wiemer2000} suggests a goodness of fit of 90\% as an acceptable threshold to select $M_c$. However, not all frequency-magnitude distributions reach the 90\% mark, in which case $M_c$ can be selected by inspection.

Figure \ref{fig:completeness} shows the goodness-of-fit values for the three seismic zones. We indicate the selected value of $M_c$ for each region in the figure. We also obtained the $b$-value in the Gutenberg-Richter law for each region following the maximum likelihood estimation \citep{Aki1965}:
% 
\begin{equation}
	b = \frac{log_{10}(e)}{\overline{M} - M_{min}} \, ,
\end{equation}
% 
where $e$ is the mathematical constant or Euler's number, $\overline{M}$ is the average magnitude, and $M_{\min}$ is the minimum magnitude in the sample. Here, the value of $M_{\min}$ is taken as the minimum completeness magnitude $M_c$ obtained or selected for each region's sub-catalog. Table \ref{tab:seismicity} shows the corresponding values of $M_c$ and $b$.

\begin{figure}%[t]
	\centering
	\includegraphics[width=0.45\textwidth]{figures/pdf/figure-04} 
	\caption{Completeness magnitude goodness-of-fit test (GFT) results and selected $M_c$ values for the three tectonic seismic regions in northern Iran as evaluated from the time-magnitude sequence of earthquakes during the 2005--2015 period. The symbols indicate the computed GFT values as function of the earthquake magnitude. The horizontal dashed line indicates the desired threshold for the GFT value at 90 percent. The vertical dashed lines and solid symbols indicate the selected completeness magnitude for each region. The color version of this figure is available only in the electronic edition.}
	\label{fig:completeness}
\end{figure}

\begin{table}
	\fontfamily{lmss}\selectfont
	\caption{Total and declustered number of earthquakes and seismic parameters $M_c$ and $b$ for the three seismic zones of northern Iran.}
	\centering\small
	\begin{tabular}{|ccccc|}
		\hline
		Region      & Total Num. & Declus.~Num. & \textit{M}c & \textit{b}-value \\ 
		            & of Events  & of Events &             &                  \\
		\hline
		Azerbaijan  &        271 &        93 &         3.5 &             1.02 \\
		Alborz      &       1262 &       794 &         2.6 &             0.77 \\
		Kopeh Dagh  &        399 &       282 &         2.6 &             0.58 \\
		\hline
	\end{tabular}
	\label{tab:seismicity}
\end{table}

\begin{figure*}[t]
	\centering
	\includegraphics[width=\textwidth]{figures/pdf/figure-05} 
	\caption{Time-magnitude sequences for the northern Iran seismic zones of Azerbaijan, Alborz and Kopeh Dagh, during the period between 1 January 2005 and 31 December 2015. The event occurrences are represented by vertical bars or sticks of length equal to the $M_w$ magnitude of each earthquake. These sequences correspond to the declustered catalog but in each case we consider only events with $M_w \geq M_c$. As shown in Fig.~\ref{fig:vg}, the dots at the top of each stick represent the nodes of the visibility graphs. The edges or links are omitted for visual convenience. Highlighted in the figure are the largest events in each sub-catalog, namely the 2012 $M_w$ 6.4 East Azerbaijan, 2010 $M_w$ 5.8 Damghan, and 2012 $M_w$ 5.4 Neyshabur earthquakes. The color version of this figure is available only in the electronic edition.}
	\label{fig:mag-time}
\end{figure*}

In addition to using $M_c$ to determine the $b$-value for each sub-catalog, we also use $M_c$ as the threshold for limiting the minimum magnitude to be considered in the construction of the visibility graph of each region. The resulting number of events for each region is included in Table \ref{tab:seismicity}. Figure \ref{fig:mag-time} shows the final time-magnitude sequences for Azerbaijan, Alborz and Kopeh Dagh. Note that in the case of Alborz and Kopeh Dagh, we consider events in the declustered catalog down to $M_w \geq 2.6$, whereas in the case of Azerbaijan all events are $M_w \geq 3.5$.

>>>>

We are interested in applying the visibility graph method to analyze the seismicity of northern Iran. This region is part of the Iranian plateau, on the Himalayan-Alpine seismic belt. It is confined by the relative movements between the Arabian, Eurasian, eastern Asia-Minor, and Indian plates; and has a long history of large magnitude ($M>7$) earthquakes that are well documented dating back to the eight century \citep[e.g.,][]{Berberian_1981_Chap}. According to the tectonic settings and geologic provinces of the plateau, the seismic activity in Iran has been categorized into different seismic zones. These vary between four to nine major seismic zones, in the more traditional models \citep[e.g.,][]{Stocklin1968, Takin1972, Berberian1976}, up to twenty to twenty-three seismotectonic provinces in the most elaborate ones \citep[e.g.,][]{Nowroozi1976, Tavakoli1999}. We adopt the model proposed by \citet{Mirzaei1998} with modifications introduced by \citet{Karimiparidari2013}. In it, Iran is divided into six seismic regions: Azerbaijan, Alborz, Kopeh Dagh, Zagros, Central-East Iran, and Makran. \myrevision{Each of these seismic regions and in general Iranian plateau are studied from different perspectives \citep[e.g.,][]{Zafarani2017testing,Sedaghati2017partially,Golara2014, Nemati2015}}. Our focus, however, is on the northern part of Iran, for which we define a region of interest between longitudes 43.5\textdegree{}E and 61.5\textdegree{}E, and latitudes 34\textdegree{}N and 40\textdegree{}N, as shown in Fig.~\ref{fig:study_region}. Although this region encloses part of the Zagros and Central-East seismic zones, we concentrate in the analysis of the seismicity of Azerbaijan, Alborz and Kopeh Dagh.

@@

\begin{figure*}[t]
	\centering
	\includegraphics[width=\textwidth]{figures/pdf/figure-03} 
	\caption{\myrevision{Instrumental} seismicity of northern Iran for the 2005--\myrevision{2016} period, considering only the events in the seismic zones of interest, namely Azerbaila, Alborz, and Kopeh Dagh. The size of symbols are proportional to the magnitude of the events, as indicated by the artificial scale shown on the right. The background shows fault lines and the shaded relief. The color version of this figure is available only in the electronic edition.}
	\label{fig:seismicity}
\end{figure*}

It follows from this description that the region of interest is one of significant seismic activity, which goes back centuries. We are, however, interested in the most recent instrumental seismicity. In particular, we focus on events recorded in the last decade, between January 2005 and December \myrevision{2016}. We focus our analysis on this magnitude-time sequence period because in a previous study we observed that the latest decade of seismic records in northern Iran offered the most complete dataset of recorded events, especially for the case of small $M < 4$ earthquakes \myrevision{\citep[][]{Khoshnevis2017}}.

We compiled a catalog of all recorded earthquakes using data downloaded from the International Institute of Earthquake Engineering and Seismology, IIEES (see the Data and Resources section). \myrevision{Instrumental catalog provided by IIEES} contained a mixture of earthquake magnitude scales, including: moment, $M_w$; local, $M_L$; body wave, $m_b$; surface wave, $M_s$; and duration $M_D$ magnitudes. \myrevision{All earthquakes of this study are reported in local $M_L$ magnitude scale. However, in order to keep consistency with other studies (e.g., completeness magnitude) we converted them to moment magnitude, $M_w$}. Magnitudes were converted using the relationships defined in \citet{Zare2014}.

\myrevision{After conducting separate studies on the whole and declustered catalog, we found that the whole catalog results are ,although slightly different, comprehensive. On the other hand declustering could add some unwanted uncertainty (in using parameters for time and distance windows) to the resultant catalog and considerably reduce the number of events in each region's catalog}. Fig.~\ref{fig:seismicity} shows the epicenter location of instrumental earthquakes for the three seismic zones of interest.

%Foreshocks and aftershocks were removed from the catalog following a Poissonian occurrence model using the declustering method of \citet{Gardner1974}, in a manner consistent with previous catalog compilations available for the region \citep[e.g.,][]{Zare2014}. Upon declustering, we divide the catalog into three sub-catalogs, each for one of the seismic zones under consideration.

We \cmmnt{then proceeded to} determine the minimum magnitude of completeness, $M_c$, for each region's sub-catalog. The value of $M_c$ is often determined using simple numerical analyses in combination with data inspection. Two common approaches are the maximum curvature (MAXC) and the goodness-of-fit test (GFT) methods \citep{Wiemer2001}. \myrevision{We compute the $M_c$ versus time using the \myrevision{MAXC} \cmmnt{GFT} method coded in ZMAP software} as introduced by \citet{Wiemer2000}.

\cmmnt{In this method, the completeness magnitude is obtained such that the catalog satisfies---at a certain acceptance threshold---the Gutenberg-Richter power low, given an extended dataset of events. This extended dataset is composed of predicted and observed events, where the predicted events are generated based on a trial minimum magnitude. The process is repeated for increasing values of the reference minimum magnitude until finding a desirable fit with the power law. \citet{Wiemer2000} suggests a goodness of fit of 90\% as an acceptable threshold to select $M_c$. However, not all frequency-magnitude distributions reach the 90\% mark, in which case $M_c$ can be selected by inspection.}

Figure \ref{fig:completeness} shows the \cmmnt{goodness-of-fit} \myrevision{$M_c\ vs\ Time$} values for the three seismic zones. We indicate the selected value of $M_c$ for each region in the figure. We also obtained the $b$-value in the Gutenberg-Richter law for each region following the maximum likelihood estimation \citep{Aki1965}:
\myrevision{
% 
\begin{equation}
	b = \frac{log_{10}(e)}{\overline{M} - (M_{min} - \frac{\Delta M_{bin}}{2})} \, ,
\end{equation}
}
% 
where $e$ is the mathematical constant or Euler's number, $\overline{M}$ is the average magnitude, and $M_{\min}$ is the minimum magnitude in the sample. Here, the value of $M_{\min}$ is taken as the minimum completeness magnitude $M_c$ obtained or selected for each region's sub-catalog. \myrevision{In this study we assumed $\Delta M_{bin} = 0.1$}. Table \ref{tab:seismicity} shows the corresponding values of $M_c$ and $b$.\myrevision{The standard deviation of the estimate of b is calculated by using \citet{Shi1982standard} formula:}
\myrevision{
\begin{equation}
	\sigma b = 2.3b^2\sqrt{\frac{\sum_{i=1}^{N}(M_i - \overline{M})^2}{N(N-1)}},
\end{equation}
where $N$ is the number of events in each catalog.
}

\cmmnt{
\begin{figure}%[t]
	\centering
	\includegraphics[width=0.45\textwidth]{figures/pdf/figure-04-rev} 
	\caption{Completeness magnitude goodness-of-fit test (GFT) results and selected $M_c$ values for the three tectonic seismic regions in northern Iran as evaluated from the time-magnitude sequence of earthquakes during the 2005--2015 period. The symbols indicate the computed GFT values as function of the earthquake magnitude. The horizontal dashed line indicates the desired threshold for the GFT value at 90 percent. The vertical dashed lines and solid symbols indicate the selected completeness magnitude for each region. The color version of this figure is available only in the electronic edition.}
	\label{fig:completeness}
\end{figure} 
}

\begin{figure}%[t]
	\centering
	\includegraphics[width=0.8\textwidth]{figures/pdf/figure-04-rev.pdf}

	\caption{\myrevision{Completeness magnitude vs time results and selected $M_c$ values for the three tectonic seismic regions in northern Iran as evaluated from the time-magnitude sequence of earthquakes during the 2005--2016 period.}}
	\label{fig:completeness}
\end{figure}

\begin{table}
	\fontfamily{lmss}\selectfont
	\caption{Total number of earthquakes \myrevision{(before and after considering completeness magnitude)} and seismic parameters $M_c$ and $b$ for the three seismic zones of northern Iran.}
	\centering\small
	\myrevision{
	\begin{tabular}{|ccccc|}
		\hline
		Region           & Total Num.  & $M_c$  & Num. of Events          & $b-value$ \\ 
		                      & of Events     &             & in Complete Catalog  &                  \\
		\hline
		Azerbaijan     &        1047     &  3.5       &        202                     &    0.85491 $\pm$ 0.0506 \\
		Alborz            &        1898     &  3.0       &        575                     &    0.81578 $\pm$ 0.0279 \\
		Kopeh Dagh  &         640      &  3.5       &        117                     &    0.78297 $\pm$ 0.0494 \\
		\hline
	\end{tabular}
	}
	\label{tab:seismicity}
\end{table}

\begin{figure*}[t]
	\centering
	\includegraphics[width=\textwidth]{figures/pdf/figure-05-rev} 
	\caption{Time-magnitude sequences for the northern Iran seismic zones of Azerbaijan, Alborz and Kopeh Dagh, during the period between 1 January 2005 and 31 December \myrevision{2016}. The event occurrences are represented by vertical bars or sticks of length equal to the $M_w$ magnitude of each earthquake. These sequences correspond to the \myrevision{whole} catalog but in each case we consider only events with $M_w \geq M_c$. As shown in Fig.~\ref{fig:vg}, the dots at the top of each stick represent the nodes of the visibility graphs. The edges or links are omitted for visual convenience. Highlighted in the figure are the largest events in each sub-catalog, namely the 2012 $M_w$ 6.4 East Azerbaijan, 2010 $M_w$ 5.8 Damghan, and 2012 $M_w$ 5.4 Neyshabur earthquakes. The color version of this figure is available only in the electronic edition.}
	\label{fig:mag-time}
\end{figure*}

In addition to using $M_c$ to determine the $b$-value for each sub-catalog, we also use $M_c$ as the threshold for limiting the minimum magnitude to be considered in the construction of the visibility graph of each region. The resulting number of events for each region is included in Table \ref{tab:seismicity}. Figure \ref{fig:mag-time} shows the final time-magnitude sequences for Azerbaijan, Alborz and Kopeh Dagh. Note that in the case of \myrevision{Azerbaijan} and Kopeh Dagh, we consider events in the \cmmnt{declustered} catalog down to $M_w \geq 3.5$, whereas in the case of \myrevision{Alborz} all events are $M_w \geq 3.0$.
